---
title: "Evaluating Causes of High Blood Pressure in Nurses"
author: "Trevor Freeland"
date: "April 25, 2018"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA, message = F, warning = F, fig.height = 4)
```

```{r Libraries}
library(tidyverse)
library(lme4)
library(pander)
library(stargazer)
library(HLMdiag)
```

##Introduction

A study by Goldstein and Shapiro (2000) collected information from 203 registered nurses in the Los Angeles area between 24 and 50 years of age on blood pressure and potential factors that contribute to hypertension. Using the information collected by Goldstein and Shapiro we are investigating what factors are significantly associated with higher systolic blood pressure. Researchers were initially interested to see if factors like family history of hypertension or menstrual phase were associated with increased blood pressure levels. Below we describe all specific variables that we are examing.

##Methods

Data from this study provided by Weiss (2005) includes about 50 observations per nurse taken repeatedly on the nurses over the course of a single day. Blood pressure was measured a half hour before the nurse's shift began, and was taken at apporximately 20 minute intervals throughout the rest of the day. At each blood pressure reading, the nurses also rated their mood on several dimensions and the activity of each subject during the 10 minutes before each reading was measured using an actigraph worn on the waist. 

Each of the variables used in our model selection process is described below:

SNUM: subject ID 
SYS: systolic blood pressure (mmHg) (**Response Variable**) 
HRT: heart rate (beats per minute) / 10 for scaling purposes for our model.
MNACT5: activity level (frequency of movements in 1-minute intervals, over a 10-minute period )  / 10 for scaling purposes for our model.
PHASE: menstrual phase (follicular—beginning with the end of menstruation and ending with ovulation, or luteal—beginning with ovulation and ending with pregnancy or menstruation)  
DAY: workday or non-workday  
POSTURE: position during blood pressure measurement—either sitting, standing, or reclining  
STR, HAP, TIR: self-ratings by each nurse of their level of stress, happiness and tiredness at the time of each blood pressure measurement on a 5-point scale, with 5 being the strongest sensation of that feeling and 1 the weakest  
AGE: age in years / 10 for scaling purposes in our model.  
FH123: coded as either NO (no family history of hypertension), YES (1 hypertensive parent), or YESYES (both parents hypertensive)  
timept: number of the measurement that day (approximately 50 for each subject)  Starting at 0 and increasing by .1 for every measurement for scaling for our model.

##Results

Before we began to fit a model needed to do some initial exploration of our data. Our first step in that process was to examine summary statistics for our data. In Table 1 below we can notice that Some of our numeric variables are on vastly different scales, especially noticing that the emotional scales only go from 1-5, whereas heart rate, activity, age and time stamps are all significant'y larger than that. We decided for our analysis and for our model we would divide all of those four variables by 10. This prevented us from noticing any 

```{r}
nurses <- read.csv("http://math.carleton.edu/Chihara/Stats345/nursebp.csv")
nurses.no.na <- na.omit(nurses)
nurses <- nurses.no.na %>% filter(SNUM != 1046)
nurses <- filter(nurses, POSTURE != "")
nurses$POSTURE <- factor(nurses$POSTURE)
nurses <- nurses %>% select(-c(DIA, time, timepass))
pander(summary(nurses))
```


```{r}
nurses$HRT <- nurses$HRT/10
nurses$MNACT5 <- nurses$MNACT5/10
nurses$AGE <- nurses$AGE/10
nurses$timept <- nurses$timept/10
pander(summary(nurses.stats))
```

##Discussion