---
title: "Work"
author: "Trevor Freeland"
date: "April 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA, message = F, warning = F, fig.height = 4)
```

```{r Libraries}
library(tidyverse)
library(lme4)
library(pander)
```

```{r Data Read}
nurses <- read.csv("http://math.carleton.edu/Chihara/Stats345/nursebp.csv")
```

SNUM: subject ID 
SYS: systolic blood pressure (mmHg) (**Response Variable**) 
DIA: diastolic blood pressure (mmHg)  
HRT: heart rate (beats per minute)  
MNACT5: activity level (frequency of movements in 1-minute intervals, over a 10-minute period )  
PHASE: menstrual phase (follicular—beginning with the end of menstruation and ending with ovulation, or luteal—beginning with ovulation and ending with pregnancy or menstruation)  
DAY: workday or non-workday  
POSTURE: position during blood pressure measurement—either sitting, standing, or reclining  
STR, HAP, TIR: self-ratings by each nurse of their level of stress, happiness and tiredness at the time of each blood pressure measurement on a 5-point scale, with 5 being the strongest sensation of that feeling and 1 the weakest  
AGE: age in years  
FH123: coded as either NO (no family history of hypertension), YES (1 hypertensive parent), or YESYES (both parents hypertensive)  
time: in minutes from midnight  
timept: number of the measurement that day (approximately 50 for each subject)  
timepass: time in minutes beginning with 0 at time point 1  

##Results

```{r}
#What data has all entries
nurses.no.na <- na.omit(nurses)
#What do the entries with NA have them in
nurses.na <- anti_join(nurses, nurses.no.na)
#If we delete observations with NA's, are we going to have too few observations for someone
temp <- nurses.no.na %>% group_by(SNUM) %>% summarize(n=n())
sort(temp$n)
#1046 only has 6 observations, going to exlcude because it might throw off time variable.
nurses <- nurses.no.na %>% filter(SNUM != 1046)
summary.stats <- summary(nurses) 
pander(summary.stats)
```

```{r}
#unconditional means model
model.means <- lmer(SYS~(1|SNUM), data = nurses)
summary(model.means)
#Intraclass Cor
cor.intra <- 65/(65+166)
#unconditional growth model
model.growth <- lmer(SYS~timept+(1|SNUM), data = nurses)
summary(model.growth)
#Change.in.resid 
resid.change <- .4/166.63 #miniscule
#First model
```