---
title: "Work"
author: "Trevor Freeland"
date: "April 25, 2018"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA, message = F, warning = F, eval = F, fig.height = 4)
```

```{r Libraries, eval = T}
library(tidyverse)
library(lme4)
library(pander)
library(HLMdiag)
library(stargazer)
library(sjPlot)
```

```{r Data Read}
nurses <- read.csv("http://math.carleton.edu/Chihara/Stats345/nursebp.csv")
```

SNUM: subject ID 
SYS: systolic blood pressure (mmHg) (**Response Variable**) 
DIA: diastolic blood pressure (mmHg)  
HRT: heart rate (beats per minute)  
MNACT5: activity level (frequency of movements in 1-minute intervals, over a 10-minute period )  
PHASE: menstrual phase (follicular—beginning with the end of menstruation and ending with ovulation, or luteal—beginning with ovulation and ending with pregnancy or menstruation)  
DAY: workday or non-workday  
POSTURE: position during blood pressure measurement—either sitting, standing, or reclining  
STR, HAP, TIR: self-ratings by each nurse of their level of stress, happiness and tiredness at the time of each blood pressure measurement on a 5-point scale, with 5 being the strongest sensation of that feeling and 1 the weakest  
AGE: age in years  
FH123: coded as either NO (no family history of hypertension), YES (1 hypertensive parent), or YESYES (both parents hypertensive)  
time: in minutes from midnight  
timept: number of the measurement that day (approximately 50 for each subject)  
timepass: time in minutes beginning with 0 at time point 1  

##Results

```{r}
#What data has all entries
nurses.no.na <- na.omit(nurses)
#What do the entries with NA have them in
nurses.na <- anti_join(nurses, nurses.no.na)
#If we delete observations with NA's, are we going to have too few observations for someone
temp <- nurses.no.na %>% group_by(SNUM) %>% summarize(n=n())
#1046 only has 6 observations, going to exlcude because it might throw off time variable.
nurses <- nurses.no.na %>% filter(SNUM != 1046)
nurses <- filter(nurses, POSTURE != "")
nurses$POSTURE <- factor(nurses$POSTURE)
summary.stats <- as.tibble(summary(nurses)) 
pander(summary(nurses))

#Should look to rescale variables. HRT and MNACT5, maybe AGE

nurses$HRT <- nurses$HRT/10
nurses$MNACT5 <- nurses$MNACT5/10
nurses$AGE <- nurses$AGE/10
nurses$timept <- nurses$timept/10
```

```{r}
#EDA
cbPalette <- c("#000000", "#E69F00")
ggplot(nurses, aes(y = SYS, x = HRT)) + geom_point() + stat_smooth(method="lm") + facet_grid(.~POSTURE) + scale_colour_manual(values=cbPalette)

ggplot(nurses, aes(y = SYS, x = timept*10, group = SNUM)) + stat_smooth(method="lm", se=FALSE, aes(color=factor(SNUM))) + theme(legend.position = "none") + xlab("Measurement Number") + ylab("Systolic Blood Pressure (mmHg)") +ggtitle("Spaghetti Plots of Nurses Blood Pressure Over the Day")
```


```{r}
#unconditional means model
model.means <- lmer(SYS~(1|SNUM), data = nurses)
summary(model.means)
#Intraclass Cor
cor.intra <- 66/(66+166)
#unconditional growth model
model.growth <- lmer(SYS~timept+(1|SNUM), data = nurses)
summary(model.growth)
#Change.in.resid 
resid.change <- .6/166.63 #miniscule change

```


```{r}
#Base model -> All of our fixed effects with full interactions and an Age and time random slope effect


model.full <- lmer(SYS~(HRT + MNACT5 + PHASE + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 +  (AGE + timept|SNUM) , data = nurses)
summary(model.full)
model1 <- lmer(SYS~(HRT + MNACT5 + PHASE + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 +  (timept|SNUM) , data = nurses)
summary(model1)

l0 <- logLik(model1)
l1 <- logLik(model.full)
D <- 2*as.numeric(l1-l0)
.5*(1-pchisq(D,3)) + .5*(1-pchisq(D,2))
#We can drop the AGE random slop

model2 <- lmer(SYS~(HRT + MNACT5 + PHASE + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 +  (1|SNUM) , data = nurses)
summary(model2)

l0 <- logLik(model2)
l1 <- logLik(model1)
D <- 2*as.numeric(l1-l0)
.5*(1-pchisq(D,2)) + .5*(1-pchisq(D,1))
#We cannot drop the random slope for time
#Sticking with model 1, random slope on time
```

```{r}
#Fixed Effcets time

model1 <- lmer(SYS~(HRT + MNACT5 + PHASE + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 +  (timept|SNUM) , data = nurses, REML = F)
summary(model1)

#Do we need any interactions?
model2 <- lmer(SYS~(HRT + MNACT5 + PHASE + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept) +  (timept|SNUM) , data = nurses, REML = F)
anova(model1,model2)
#Yeah there is something in there that we need.

model3 <- lmer(SYS~(PHASE + MNACT5 + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 + (HRT*POSTURE) +  (timept|SNUM) , data = nurses, REML = F)
anova(model3, model1)
summary(model3)

#HRT only had a few sig interactions

model4 <- lmer(SYS~(PHASE*MNACT5) + (MNACT5 + DAY + POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 + (HRT*POSTURE) +  (timept|SNUM) , data = nurses, REML = F)
anova(model3, model4)
summary(model4)

#PHASE only had 1 sig intercation.

model5 <- lmer(SYS~(MNACT5 +  POSTURE + STR + HAP + TIR + AGE + FH123 + timept)^2 + (DAY * AGE) + (HRT + DAY)*POSTURE + (PHASE*MNACT5) + (timept|SNUM) , data = nurses, REML = F)
anova(model4, model5)
summary(model5)

#Day needs interaction with age and posture

model6 <- lmer(SYS~(HAP+MNACT5+POSTURE+TIR+AGE+FH123+timept)^2+(DAY*AGE)+(HRT+STR+DAY)*POSTURE + (PHASE*MNACT5) + (timept|SNUM) , data = nurses, REML = F)
anova(model6, model5)
summary(model6)

#STR needed Posture

model7 <- lmer(SYS~(MNACT5+POSTURE+TIR+AGE+FH123+timept)^2+(DAY*AGE)+(HRT+HAP+STR+DAY)*POSTURE + (PHASE)*MNACT5 + (timept|SNUM) , data = nurses, REML = F)
anova(model6, model7)
summary(model7)

#HAP needs Posture

model8 <- lmer(SYS~(AGE+FH123+MNACT5+POSTURE+timept)^2+(DAY*AGE)+(HRT+HAP+STR+DAY)*POSTURE + (PHASE)*MNACT5 + (timept|SNUM) , data = nurses, REML = F)
anova(model8, model7)
summary(model8)

#TIR unneccesary

model9 <- lmer(SYS~(MNACT5+POSTURE+timept)^2+(DAY*AGE)+(HRT+HAP+STR+DAY)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses, REML = F)
anova(model8, model9)
summary(model9)

#FH123 only needs time interaction.

model10 <- lmer(SYS~(MNACT5+POSTURE)^2+(DAY*AGE)+(HRT+HAP+STR+DAY)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses, REML = F)
anova(model10, model9)
summary(model10)

#timept only needs interaction with FH123

model11 <- lmer(SYS~(DAY*AGE)+(HRT+HAP+STR+DAY+MNACT5)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses, REML = F)
anova(model10, model11)
summary(model11)

#I think that is the model I am going to stick with
```

```{r}
# Diagnostics
model11 <- lmer(SYS~(DAY*AGE)+(HRT+HAP+STR+DAY+MNACT5)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses, REML = F)
x2 <- HLMresid(model11, level = 1, standardize = TRUE)
plot(fitted(model11), x2, ylab = "Conditional Residual")
#There appears to be some potential outliers with high fitted values
```

```{r}
x <- HLMresid(model11, level="marginal",standardize=TRUE)
plot(x~fitted(model11), ylab = "Marginal Residual")
abline(h=3)
abline(h=-3)
#There appears to be some potential outliers
```

```{r}
cd <- cooks.distance(model11)
dotplot_diag(cd,cutoff="internal", name="cooks.distance")
index1 <- which(cd>.004)
```

```{r}
lev <- leverage(model11, level = 1)
dotplot_diag(lev[,1],cutoff="internal",name="leverage")
index2 <- lev[,1] > .1
nurses[index2,]
```

```{r}
#trying to re run model taking out high cooks distance points (some of the high leverage points were also large cooks distance points so I wanted to take out the cooks distance points first to see any effect)
nurses.test <- nurses[-index1,]
model.test <- lmer(SYS~(DAY*AGE)+(HRT+HAP+STR+DAY+MNACT5)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses.test, REML = F)
summary(model.test)
summary(model11)

#Taking away the large cooks distance values greatly change the model. So we should evaluate the model without them. 

model.a <- lmer(SYS~(DAY*AGE)+(HRT+HAP+MNACT5)*POSTURE +DAY + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses.test, REML = F)
anova(model.a, model.test)
summary(model.a)

#took out STR all together and Day interactions with Posture

model.b <- lmer(SYS~(DAY*AGE)+HRT+(HAP+MNACT5)*POSTURE +DAY + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses.test, REML = F)
anova(model.a, model.b)
summary(model.b)

#Can't take out HRT:Posture interactions so model.a looks like new model. Lets re-check diagnostics
```

```{r}
# Diagnostics
x2 <- HLMresid(model.a, level = 1, standardize = TRUE)
plot(fitted(model.a), x2, ylab = "Conditional Residual")
#This plot looks more or less the same as the last one. Maybe slightly fewer outliers?
```

```{r}
x <- HLMresid(model.a, level="marginal",standardize=TRUE)
plot(x~fitted(model.a), ylab = "Marginal Residual")
abline(h=3)
abline(h=-3)
#There still appears to be some potential outliers, again maybe a few less.
```

```{r}
cd <- cooks.distance(model.a)
dotplot_diag(cd,cutoff="internal", name="cooks.distance")
index1 <- which(cd>.001)
```

```{r}
lev <- leverage(model.a, level = 1)
dotplot_diag(lev[,1],cutoff="internal",name="leverage")
index2 <- lev[,1] > .1
nurses.test[index2,]
nurses.test[index1,]

nurses.test2 <- nurses.test[-index1,]

model.a <- lmer(SYS~(DAY*AGE)+(HRT+HAP+MNACT5)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses.test, REML = F)
model.test.a <- lmer(SYS~(DAY*AGE)+(HRT+HAP+MNACT5)*POSTURE + (PHASE)*MNACT5 + (FH123)*timept + (timept|SNUM) , data = nurses.test, REML = F)
anova(model.a,model.test.a)
summary(model.a)
#our estimates don't change significantly after excluding this data, so we will keep it in the model. 
#final model is model.a with nurses.test as the final data
```

Model:

Level 1: $BloodPressure_{ij} = a_i + b_0(HeartRate) + c_0(Happiness) + d_i(Activity) + e_0(Stand) + f_0(Sit)$
$+ g_i(time) + h_0(HeartRate:Stand) + h_1(HeartRate:Sit) + l_0(Happiness:Sit) + l_1(Happiness:Stand) +$
$k_0(Activity:Sit) + k_1(Activity:Sit) + \epsilon_{ij}$

Level 2: $a_i = a_0 + a_1(WorkDay) + a_2(Age) + a_3(Age:WorkDay) + a_4(PhaseL) + a_5(OneHyperPar) + a_6(TwoHyperPar) + \mu_i$

$d_i = d_0 + d_1(PhaseL)$

$g_i = g_0 + g_1(OneHyperPar) + g_2(TwoHyperPar) + \omega_i$

$\mu_i, \omega_i$~BivariateNormal means = [0,0] \[
   \Sigma = 
  \left[ {\begin{array}{cc}
   \sigma_1^2 & \sigma_{12} \\
   \sigma_{12}  & \sigma_2^2  \\
  \end{array} } \right]
\]
